Lesson 3
========================================================
# Read in data
```{r}
#use sep to indicate file is tab seperated
pf <- read.csv('pseudo_facebook.tsv', sep = '\t')
names(pf)
```

***

### Histogram of Users' Birthdays
Notes:

```{r Histogram of Users\' Birthdays}
#install.packages('ggplot2')
library(ggplot2)

qplot(x=dob_day, data=pf) +
  scale_x_continuous(breaks = 1:31)
```

***

#### What are some things that you notice about this histogram?
Response:
# A large spike on the first day. This is the default so a lot of people probably didnt bother changing to correct date. The dip on the 31st is to be expected as not all months have 31 days in them

***
### Faceting
Notes:

```{r Faceting}
# Get a plot for each of the 12 months. ncol specifies having 3 columns of histograms
qplot(x=dob_day, data = pf) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~dob_month, ncol =3)
```

#### Let’s take another look at our plot. What stands out to you here?
Response:
# The spike for the first date of a month only happens in January. Calendar must be preset to Jan 1st and people arent bothered changing it

***

#### What code would you enter to create a histogram of friend counts?

```{r Friend Count}
qplot(x=friend_count, data=pf)
```


### Limiting the Axes
Notes: The bulk of out data is below the 1000 friends point so we'll focus on that using xlim or use: +
  scale_x_continuous(limits = c(0, 1000))

```{r Limiting the Axes}
qplot(x=friend_count, data=pf, xlim = c(0,1000)) 
  
```

***

### Adjusting the Bin Width
Notes:
breaks = seq(0, 1000, 50) says go from 0 to 1000 in steps of 50

### Faceting Friend Count
```{r Faceting Friend Count}
# create a facet the histogram by gender?
qplot(x = friend_count, data = pf, binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Omitting NA Values
Notes: could use data=na.omit(pf) but this will omit any rows that have na for any of their values and not just gender

```{r Omitting NA Values}
qplot(x = friend_count, data = subset(pf, !is.na(gender)), binwidth = 10) +
  scale_x_continuous(limits = c(0, 1000),
                     breaks = seq(0, 1000, 50)) +
  facet_wrap(~gender)
```

***

### Statistics 'by' Gender
Notes:

```{r Statistics \'by\' Gender}
table(pf$gender)
by(pf$friend_count, pf$gender, summary)
```


#### Why would the median be a better measure than the mean?
Response: Median is a more robust statistic. A few people with huge friend counts drag the mean upwards which isnt representative of most users. Median is not affected by these extreme values

***

### Tenure
Notes:

```{r Tenure}
qplot(x=tenure, data = pf, binwidth=30, 
      color=I('black'), fill=I('#099DD9'))
```

***

#### How would you create a histogram of tenure by year?

```{r Tenure Histogram by Year}
qplot(x=tenure/365, data = pf, binwidth=0.25, 
      color=I('black'), fill=I('#F79420')  )
```

***

### Labeling Plots
Notes:

```{r Labeling Plots}

qplot(x=tenure/365, data = pf, binwidth=0.25, 
      xlab ='Number of years using Facebook',
      ylab='Number of users in sample',
      color=I('black'), fill=I('#F79420') )
```

***

### User Ages
Notes:
Running summary(pf$age) will give min and max ages
seq(0,113,5) goes from 0 to 113 in units of 5
```{r User Ages}
qplot(x=age, data=pf, binwidth=1, xlab = "age", 
      ylab='Number of users in sample',
      color=I('black'), fill=I('#5760AB')) +
      scale_x_continuous(breaks = seq(0,113,5))  
```

***

### Transforming Data
Notes:
```{r}
#qplot(x=friend_count, data =pf)
summary(pf$friend_count)
#getting log10 of 0 is undefined so the min - where people have 0 friends
#would yield a result of negative infinity. In order to avoid this we add # 1 to the friend count
summary(log10(pf$friend_count +1))
```

```{r Transfroming Data Quiz}
#install.packages('gridExtra')
library(gridExtra)

# define individual plots
p1 <- qplot( x=friend_count, data = pf) + labs(x = 'number of friends', y = 'count') + xlim(0,2000)
p2 <- qplot( x=(log10(friend_count +1)), data = pf) 
p3 <- qplot( x=(sqrt(friend_count)), data = pf) 

# arrange plots in grid
grid.arrange(p1, p2, p3, ncol=1)
```

***

### Add a Scaling Layer
Notes:

```{r Add a Scaling Layer}
#Alternative method for above
p1 <- ggplot(aes(x=friend_count), data=pf) + geom_histogram()
p2<- p1 + scale_x_log10()
p3<- p1 + scale_x_sqrt()

# arrange plots in grid
grid.arrange(p1, p2, p3, ncol=1)
#Keep in mind this method has the same label for the x axis in all plots but it would be more helpful to know when log10 or sqrt was used in the labels
```

***


### Frequency Polygons
Comparing male and female friend counts in an overlay with a *frequency polygon* rather than two histograms side by side.
Set color = gender so we get two different coloured lines for each of the genders.
Change y axis to show prportions instead of counts:
y = ..count../sum(..count..)


```{r Frequency Polygons}
ggplot(aes(x = friend_count, y = ..count../sum(..count..)),
       data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color = gender), binwidth=10) +
  scale_x_continuous(limits = c(0, 1000), breaks = seq(0, 1000, 50)) +
  xlab('Friend Count') +
  ylab('Proportion of users with that friend count')
```

***

### Likes on the Web
Notes:

```{r Likes on the Web}
ggplot(aes(x = www_likes), data = subset(pf, !is.na(gender))) +
  geom_freqpoly(aes(color=gender)) +
  scale_x_log10() +
  xlab("Log 10 Number of likes") 
```


***
```{r}
#What's the www_like count for males?
#Which gender has more likes?

#Want to spli the like by gender and perform the sum function:
by(pf$www_likes, pf$gender, sum)


```


### Box Plots
Notes:
With boxplots the continuous variable will be displayed on the y axis and the grouping/categorical variable will be on the x axis

* The boxes cover the middle 50% of values or what's called the IQR 
* The dots represent the outliers - there a re a lot present in this graph
* Outliers are points that lie outside of 1.5 times the IQR from the median
* The horizontal line inside the boxes is the median

```{r Box Plots}
ggplot(aes(x = gender, y=friend_count), 
       data = subset(pf, !is.na(gender))) +
  geom_boxplot()
```

#### Adjust the code to focus on users who have friend counts between 0 and 1000.
* When you use the *ylim* or *scale_y_continuous* paramater you're actually removing points from the calculations.
* A better way to do this is to use *coord_cartesian*
```{r}
ggplot(aes(x = gender, y=friend_count), 
       data = subset(pf, !is.na(gender))) +
  geom_boxplot() +
  coord_cartesian(ylim=c(0,1000))

```

***

### Box Plots, Quartiles, and Friendships
Notes:
```{r Box Plots, Quartiles, and Friendships}
ggplot(aes(x = gender, y=friendships_initiated), 
       data = subset(pf, !is.na(gender))) +
  geom_boxplot() +
  coord_cartesian(ylim=c(0,150))
by(pf$friendships_initiated, pf$gender, summary)
```

***

### Getting Logical
Notes:

```{r Getting Logical}
summary(pf$mobile_likes)

#See if someone has actually checked in on a mobile device rather than the number of times:
summary(pf$mobile_likes>0)
#create our own variable to track whather a user had logged in on mbile
mobile_check_in <- NA
#Assign value of 1 if user has checked in on mobile and 0 otherwise:
pf$mobile_check_in <- ifelse(pf$mobile_likes>0, 1, 0)
#convert to factor variable
pf$mobile_check_in <- factor(pf$mobile_check_in)
summary(pf$mobile_check_in)
#Find % of users signed in on mobile:
percent_mobile_check_in = sum(pf$mobile_check_in==1)/length(pf$mobile_check_in)
percent_mobile_check_in
```


***
```{r}
#install.packages('knitr', dependencies = T)
library(knitr)
setwd("/Users/Sian/Intro_data_analysis/EDA/Lesson 3")
```

Click **KnitHTML** to see all of your hard work and to have an html
page of this lesson, your answers, and your notes!