Demystifying R Part 2
========================================================

1. The summary() function gives us an idea of the values a variable can take on.
```{r}
summary(mtcars)
```

The str() and summary() functions are helpful commands when working with a new data set.
The str() function gives us the variable names and their types.


2. 
Subset the data so that you create a new data frame called efficient that contains
cars that get 23 or more mpg (miles per gallon). 
```{r}
efficient <- subset(mtcars, mpg >= 23)
```

3. How many cars get more than 23 mpg?
```{r}
dim(efficient)
head(efficient, 10)
```

4. We can also use logical operators to find out which car(s) get greater
than 30 miles per gallon (mpg) and have more than 100 raw horsepower.
```{r}
subset(mtcars, mpg > 30 & hp > 100)
```

Note: R uses one single & for the logical operator AND. It also uses one | for the logical operator OR.

6. Print the cars that have a 1/4 mile time (qsec) less than or equal to
16.90 seconds to the console.
```{r}
subset(mtcars, qsec<= 16.9)
```

7. Save the subset of cars that weigh under 2000 pounds (weight is measured in lb/1000)
to a variable called lightCars. Print the numbers of cars and the subset to the console.
```{r}
lightCars <- subset(mtcars, wt<2.0 )
dim(lightCars)
head(lightCars)
```

8. You can also create new variables in a data frame:
```{r}
mtcars$year <- 1974
```
This assigns the year 1974 to all of the entries.

To drop a variable:
```{r}
mtcars <- subset(mtcars, select = -year)
```


9. What do you think this code does? Run it to find out.
```{r}
mtcars$year <- c(1973, 1974)
head(mtcars)
```


Drop the year variable from the data set.
```{r}
mtcars <- subset(mtcars, select = -year)
```


10. If else statements
```{r}
mtcars$wt
cond <- mtcars$wt < 3
cond #prints true if weight<3 and fa;se otherwise
mtcars$weight_class <- ifelse(cond, 'light', 'average')
mtcars$weight_class #if weight<3 true weight_class= light else weight_class=average
cond <- mtcars$wt > 3.5
mtcars$weight_class <- ifelse(cond, 'heavy', mtcars$weight_class)
mtcars$weight_class
```

You have some variables in your workspace or environment like 'cond' and
efficient. You want to be careful that you don't bring in too much data
into R at once since R will hold all the data in working memory. We have
nothing to worry about here, but let's delete those variables from the
work space.

```{r}
rm(cond)
rm(efficient)
```

Knit the HTNL document:
```{r}
#install.packages('knitr', dependencies = T)
#library(knitr)
```

Which car(s) have an mpg (miles per gallon) greater than or equal to 30
OR hp (horsepower) less than 60? yCreate an R chunk of code to answer the question.
```{r}
subset(mtcars, mpg >= 30 | hp < 60)
```

